 

### 2021-06-30 vertical-thermal-review.Rmd
attempted to organize Tleaf section
I added a seperate para keeping the one that's already there, to see if this flows better?

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating macro and microclimate [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017] by buffering extreme temperatures along with other macroclimatic conditions in the understory. 
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsWithincanopyVariationRate2004a].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**. Shown are  height profiles in growing season (a) leaf area density, (b) proportion light incidence, (c) proportion sun leaves, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

@@

**The buffering patterns described above vary geographically, seasonally, and across forests types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, *Defrenne et al. 2021*].**
The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture [**REF**], and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]. (*Nidhi, please make sure the references are appropriate here-- I rearranged the sentence structure.*)
Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; @fauset_differences_2018; **double check these refs--at least one didn't belong**].

@@

>>>>

**Rising temperatures are likely to have differential impacts on trees of different size and canopy position, in large part because small understory trees exist in a microenvironment that is substantially buffered by the more exposed canopy trees [@zellweger_seasonal_2019; @davisMicroclimaticBufferingForests2019].**
Forests are vertically and horizontally stratified, with trees of different sizes through various successional stages. 
Overstory trees form canopies that play a crucial role in moderating forest macroclimates [@ozanneBiodiversityMeetsAtmosphere2003; @nakamuraForestsTheirCanopies2017], as well as microclimates by buffering understories from extreme temperatures and other macroclimatic conditions [@zellweger_seasonal_2019].
This creates a vertical stratification of biophysical environmental conditions such as temperature, light, wind, humidity and CO~2~ that influences leaf traits, thermoregulation and metabolism along the gradient, with implications for whole plant performance [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016].
Despite the fact that this vertical gradient inevitably shapes nearly every aspect of plant metabolism, demography, and ecology, we lack comprehensive understanding of these gradients [but see @niinemetsWithincanopyVariationRate2004a].
Importantly, this limits our ability to understand how warming temperatures will affect leaf-level metabolism, whole-plant performance, and, in turn, forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions.

@@

![**Figure `r fig_NEON_vertical`. Vertical gradients in the biophysical environment for six sites in the National Ecological Observatory Network (NEON)**. Shown are  height profiles in growing season (a) leaf area density, (b) proportion sun leaves, (c) proportion light incidence, and July mean ± 1 standard deviation for (d) maximum photosyntehtically active ratiation (PAR), (e) maximum wind speed, (f) minimum humidity, (g) maximum $T_{air}$, and (h) maximum biological temperature, $T_{bio}$. Sites include a tropical montane broadleaf evergreen forest (Pu‘u Maka‘ala Natural Area Reserve, Hawaii; PUUM), a subtropical longleaf pine savanna (Ordway-Swisher Biological Station, FL; OSBS), two temperate broadleaf forests (Smithsonian Conservation Biology Institute, VA, SCBI; Smithsonian Environmental Research Center, MD, SERC), and a mixed northern hardwood and coniferous forest (Harvard Forest, MA; HARV). Vertical profiles in micrometeorological variables (d-h) at all NEON sites are shown in Fig. S1.](NEON_height_profiles/figures/profile_all.png)

@@

**The buffering patterns described above vary geographically, seasonally, and across forests types [Figs. `r fig_NEON_vertical`, `r SIfig_NEON_all`, *Defrenne et al. 2021*].**
The strength of buffering increases with canopy density -- for example, across gradients of elevation [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @juckerCanopyStructureTopography2018], moisture [**REF**], and proximity to water, for example in riparian forests [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]. (*Nidhi, please make sure the references are appropriate here-- I rearranged the sentence structure.*)
Seasonally, canopy buffering capacity is greater in wet than dry seasons, presumably because leaf area is greater in wet seasons [@davisMicroclimaticBufferingForests2019; @rey-sanchez_spatial_2016; @vonarxSpatiotemporalEffectsForest2012a; **double check these refs--at least one didn't belong**].

@@

**attempted to oraganize**

~Solar and longwave radiation can dramatically elevate $T_{leaf}$ well above $T_{air}$ (Fig. `r fig_leaf_T`a) because the heat absorbed through high irradiance can be greater than heat lost through sensible heat (via $ga$ ie. boundary layer conductance) and latent heat loss ( via $gs$ ie. Stomatal conductance and $E$ ie evaporation).

$Ga$ and $Gs$ are dominant factors in $T_{leaf}$ flux. $Ga$ increases with wind speed which also in turn increases $gs$, thereby reduces $T_{leaf}$-$T_{air}$ (Fig. `r fig_leaf_T`b). Therefore, high windspeeds in upper canopies such as in NEON sites with windspeeds ~4 ms (Fig. `r fig_NEON_vertical`) are capable of lowering $T_{leaf}-T_{air}$ that are exposed to high radiation loads, upto ~ 4$^\circ$C (Fig. `r fig_leaf_T`b)[Bonan 2008]. However, low wind speeds typically decouple leaves from $T_{air}$, more so when solar radiation is high and $gs$ is low.

Latent heat loss through $gs$, cools the leaf and decreases $T_{leaf}$-$T_{air}$. However in conditions of low soil moisture during drought (Fig. `r fig_leaf_T`), greater temperature than $gs$ thermal optimum, and air dryness causing an increase in VPD, $gs$ becomes limited, thereby increasing $T_{leaf}$-$T_{air}$. Limitation to $gs$ can also exaggerate effects of biophysical factors that contribute to energy inputs such as, solar radiation, humidity and $T_{air}$ (Fig. `r fig_leaf_T`).

Leaf size and physiology influence $ga$, $gs$, leaf thermal capacitance (capacity to hold heat), and therefore also $T_{leaf}$-$T_{air}$. Smaller leaves have thinner boundary layer that enables coupling with $T_{air}$ through higher $ga$ compared to large leaves that can be more decoupled from $T_{air}$ (Fig. `r fig_leaf_T`a). However, greater $gs$ in large leaves under low raditation, can be cooler than small leaves. Sun leaves, such as in upper canopies, tend to have greater $gs$, due to greater stomatal density [Table 1] and lower thermal capacitance due to lower water potential than shade leaves. But exposure to high irradiance in upper canopies which is a dominant source of energy input increases $T_{leaf}$-$T_{leaf}$ more than shade leaves that are shielded from biophysical conditions. However, because shade leaves have higher thermal capacitance and comparitely lower $gs$, $T_{leaf}$-$T_{air}$ can increase with slight increase in radiation (Fig. `r fig_leaf_T`a).~
